Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-15 appendix.tex
about to edit numerical

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low-frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

\subsubsection{Instability with vertical shear}
For large $\khat^2$, the low-frequency branch may be obtained by balancing
the last two terms of Eq. \ref{simple_growth} . Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right).\label{simple_growth2}
\end{align}
This signifies instability for $L\geq1$ --- the VSI ---   
since we can choose the sign of the square root to make $\hat{\nu} =
\imag\hat{\sigma}>0$.  These are the low frequency unstable modes observed
in our numerical calculations in \S\ref{vertiso_pertiso}.

>>>>

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

\subsubsection{Instability with vertical shear}
The VSI corresponds to unstable inertial waves. This is readily
obtained for large $\khat^2$ by balancing
the last two terms of Eq. \ref{simple_growth}  to obtain the low
frequency branch. Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right).\label{simple_growth2}
\end{align}
This signifies instability for $L\geq1$ since we can choose the sign
of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.  These
are the low frequency unstable modes seen in
Fig. \ref{compare_modes_iso_kx10} and
Fig. \ref{compare_modes_iso_kx30} , for which $\nu\propto\omega$.

### 2015-03-15 application.tex
about to edit numerical

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical 
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, we either only find the fundamental mode, or that it is the
most unstable, except for the appearence of some surface modes when considering large
$\khat\gtrsim 50$. However, surface modes are dependent on
boundary conditions and disturbances with large $\khat$ are more likely
subject to viscous decay, as discussed above. The dominance of the
fundamental mode is also observed in hydrodynamic
simulations of \cite{stoll14} which treat radiative cooling properly.

% The dominance of the
% fundamental mode is consistent with our analytical discussion
% (\S\ref{iso_vsi_beta_crit}) and previous numerical results (\S\ref{therm_relax_eff}), which indicate  
% higher order modes are more easily stabilized by finite thermal
% relaxation.

% In fact, for $\khat \lesssim 30$ we either only find the
% fundamental mode, or that it is the most unstable, and the associated
% growth rates are insensitive to $z_\mathrm{max}$. This is consistent with our
% analytical discussion (\S\ref{iso_vsi_beta_crit}) and previous
% numerical results (\S\ref{therm_relax_eff}), which indicate  
% higher order modes are more easily stabilized by finite thermal
% relaxation.   
% However, for large $\khat$ and/or $z_\mathrm{max}$, we also found
% surface modes with shorter growth times than those shown in
% Fig. \ref{mmsn_overall}. Such modes may not be robust because    
% they are affected by boundary conditions (\S\ref{surf_comment}) and/or
% viscosity in a real disk.

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. It is necessary to
consider smaller scales as $r$ decreases as the disk generally becomes
optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
midplane thermal relaxation timescales for the MMSN ( 
Fig. \ref{mmsn_bcrit_bcool}, left middle panel). For example, for 
$\khat=10$ we find $\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim
10$AU at $z=0$; and Fig. \ref{mmsn_overall} show stabilization inside
$\sim 8$AU where $t_\mathrm{grow}\to\infty$.

@@

\subsection{Numerical examples} 
We compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

 \begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
     and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding thin horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height. 
Accordingly we expect, and find, the VSI 
operates at $r=50$AU. The inner disk ($r=5$AU) is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column due to the vertically
global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests that the instability can operate 
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU: %  for which  $\beta <
% \beta_\mathrm{crit}  \forall z$: 
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. 
    \label{mmsn_eigenvz}}
\end{figure}

>>>>

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale given by
Eq. \ref{beta_mmsn_simp}.  Fig. \ref{beta_compare} show examples of
the thermal timescales at $r=5$AU and $r=50$AU for a
perturbation wavenumber $\khat=30$. We also plot the 
corresponding critical thermal timescales evaluated for the MMSN
(Eq. \ref{bcrit_mmsn}).  
 \begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
     and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height.
Accordingly we expect, and find, the VSI operates at $r=50$AU. The
inner disk ($r=5$AU) is optically thick near the midplane but becomes
optically thin away from it, such that $\beta < \beta_\mathrm{crit}$
for $|z|\gtrsim1.25H$. In fact, we also find the VSI at $r=5$AU, but
with much reduced growth rates (when scaled by the local rotation).

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$.  In fact, we either only find the fundamental mode, or
that it is the most unstable, except for the appearence of some
surface modes towards the inner disk when considering large
wavenumbers $\khat\gtrsim 50$. However,
surface modes are affected by boundary conditions and disturbances
with large $\khat$ are more likely subject to viscous decay, as
discussed above. The dominance of the fundamental mode is also
observed in hydrodynamic simulations of \cite{stoll14} which treat
radiative cooling properly.

We can therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. It is necessary to
consider smaller scales with decreasing $r$ as the disk generally becomes
optically thick, and growth times rise rapidly towards the inner disk.

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane. %because of fundamental mode

We thus find Fig. \ref{mmsn_overall} to be roughly consistent with
comparing the midplane thermal relaxation timescales to the critical
thermal timescale. For example, 
Fig. \ref{mmsn_bcrit_bcool} show that $\beta(\khat=10,z=0) \gtrsim
\beta_\mathrm{crit}$ for $r\lesssim 10$AU in the MMSN (left middle panel); and Fig. \ref{mmsn_overall}
show that modes with $\khat=10$ are stabilized inside $\sim 8$AU (where
$t_\mathrm{grow}\to\infty$). Our numerical results suggest the VSI
is efficient at radii of few tens of AU in the MMSN, consistent with
estimates made by \cite{nelson13}.

@@

### 2015-03-15 linear_adia.tex
about to edit numerical

% The thin disk approximation allows us to develop explicit solutions,
% but some general results may be established without it (see,
% e.g. Appendix \ref{global_corr} and \S\ref{iso_pert}).

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\chi = 1$. This
case has been investigated numerically by \cite{nelson13} and 
\cite{mcnally14}, the latter of which employed the VGSB framework.

In Appendix \ref{iso_discuss}, we give an analytical discussion of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop analytic
solutions in the thin disk limit. The methods used in Appendix
\ref{iso_discuss} and results obtained therein motivates the solution
procedure employed in the following sections.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$ % , for which the perturbations may be 
% considered nearly-isothermal,
we expect the solution to only differ slightly from a case with
$\beta\equiv 0$. We thus perturb a solution for $\beta=0$ to see the
effect of thermal relaxation.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to dominate in numerical calculations with 
increasing $\beta$, as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).  %Their importance is highlighted

@@

Inserting Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}
and keeping only first order terms yield the requirements, after
balancing constants and coefficients of $\zhat^2$,

\begin{align}
  0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
  \left(\hat{\sigma}^2 -1\right)\delta\chi, \\
  0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
    q \khat\right)\delta\chi.
\end{align}

Using the expressions for $\delta\chi$ and $\hat{\sigma}$ we can solve
for $\delta\hat{\sigma}$. We find imaginary part of $\delta\hat{\sigma}$
is 
% This gives 
% \begin{align}\label{nearly_iso_dsig}
%   \delta \hat{\sigma} =
%   -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
%       q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
% \end{align}
% The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by small finite cooling is ineffective at very large or
very small scales.

A finite thermal relaxation time introduces buoyancy
forces, which is stabilizing for sub-adiabatically stratified disks
($\gamma > 1$). The $\zhat^2$ dependence in $\delta v_{z1}$ then makes
sense because it becomes significant for large $|\zhat|$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero.

\subsection{Explicit solutions and dispersion relation}\label{disp_relax}
%To obtain a dispersion relation similar to that for 
%isothermal perturbations (Eq. \ref{sig2_iso}), 
We now solve Eq. \ref{nearly_iso_explicit} explicitly. We first write  
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\alpha -
    1/2\right)\zhat^2}$ to remain finite as $|\zhat|\to\infty$. 
Then assuming $g(\zhat)$ is a polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align} 
%by considering the case $q=0$ and $\beta\to\infty$ as examined in
%\cite{lubow93}.

@@

% However, for the fundamental mode
% ($M=0$) some analytic progress can be made, which we consider next.

\subsection{Upper limit on the thermal relaxation timescale for the
  VSI}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta$ that allows the VSI.  
%for modes with $\khat^2\gg M+1$.  
% which, as we will find, correspond to low frequency
% modes. 
Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
stability and $\khat^2\gg 1$. We consider $M=0$ (the 
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ modes such that $|\hat{\omega}\khat|$ is $O(1)$. We 
also assume $|q|$ and $\beta$ are $O(1)$ or smaller. For a thin disk
($\epsilon \ll 1$) we have 
\begin{align}
 0 = &M(M+1)\left(1-\epsilon^2q^2\khat^2\right) \notag\\ 
 &+ \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
   -  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$ (despite $\khat^2\gg 1$), 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$ Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
For modes satisfying the above assumptions (or $M=0$) so that the
second term in square brackets is small, we have 
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without limiting the magnitude of $|\epsilon q   
\khat|$.  We will confirm Eq. \ref{iso_vsi_cond} numerically in
\S\ref{bcrit_num_test}.

Eq. \ref{iso_vsi_cond} suggests that as the thermal timescale is
increased from zero, higher order modes with $M>0$
are stabilized before the fundamental mode. We indeed observe this
trend in numerical calculations, despite adopting somewhat restrictive
assumptions in order to arrive at the simple expression 
Eq. \ref{bcrit_gen} for $M>0$.

>>>>

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\gamma=1$ so 
that $\chi = 1$. This case has been investigated numerically
by \cite{nelson13} and \cite{mcnally14}.

In Appendix \ref{iso_discuss}, we give an analytical discussion of the
linear problem for isothermal perturbations. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop 
solutions in the thin disk limit. The 
methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of thermal relaxation.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to dominate in numerical calculations with 
increasing $\beta$ (\S\ref{therm_relax_eff}), as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).

@@

We insert Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso_explicit}, keeping only first order terms, and
solve for $\delta\hat{\sigma}$ using the above xpressions for $\delta\chi$
and $\hat{\sigma}$. We find imaginary part of $\delta\hat{\sigma}$
is 
% yield the requirements, after
% balancing constants and coefficients of $\zhat^2$, 
% \begin{align}
%   0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
%   \left(\hat{\sigma}^2 -1\right)\delta\chi, \\
%   0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
%     q \khat\right)\delta\chi.
% \end{align}
% This gives 
% \begin{align}\label{nearly_iso_dsig}
%   \delta \hat{\sigma} =
%   -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
%       q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
% \end{align}
% The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization, since $\gamma>1$. 
% Interestingly, for both
% $\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
% stabilization by small finite cooling is ineffective at very large or
% very small scales.

A finite thermal relaxation time allow buoyancy
forces to stabilize vertical motions in sub-adiabatically stratified
disks. The $\zhat^2$ dependence in $\delta v_{z1}$ makes  
sense because it becomes significant at large $|\zhat|$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero (since $N_z^2\propto z^2$ for a thin disk, see
Eq. \ref{nzsq_def}).

\subsection{Explicit solutions and dispersion relation}\label{disp_relax}
We now solve Eq. \ref{nearly_iso_explicit} explicitly. We first write  
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\alpha -
    1/2\right)\zhat^2}$ to remain finite as $|\zhat|\to\infty$. 
Then assuming $g(\zhat)$ is a polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align}

@@

\subsection{Upper limit on the thermal relaxation timescale for the
  VSI}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta$ that allows the VSI.  
Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
stability and $\khat^2\gg 1$. We consider $M=0$ (the 
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$. We 
also assume $|q|$ and $\beta$ are $O(1)$ or smaller. For a thin disk
($\epsilon \ll 1$) we have 
\begin{align}
 0 = &M(M+1)\left(1-\epsilon^2q^2\khat^2\right) \notag\\ 
 &+ \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
   -  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$ Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since the second term in square brackets is small by assumption, we have 
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without assuming $|\epsilon q\khat|\lesssim 1$.  We will confirm Eq. \ref{iso_vsi_cond} numerically in 
\S\ref{bcrit_num_test}.

Eq. \ref{iso_vsi_cond} suggests that as the thermal timescale $\beta$
is increased from zero, higher order modes with $M>0$
will be stabilized before the fundamental mode. We indeed observe this
trend in numerical calculations (\S\ref{therm_relax_eff}), despite
adopting somewhat restrictive assumptions for $M>0$ in order to arrive
at the simple expression Eq. \ref{bcrit_gen}.
